---
title: "BMT study day 14"
author: "Dawud"
date: "01/30/2022"
output: html_document
---

```{css, echo=FALSE}
.watch-out {
  background-color: rgb(240, 250, 250);
  border: 1px solid green;
  font-family:Consolas;
  
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(class.source = "watch-out")
```


## Load packages

```{r message=F, warning=F}
# Packages from CRAN:
if(!require(tidyverse)){install.packages("tidyverse")}
if(!require(ggpubr)){install.packages("ggpubr")}
if(!require(vegan)){install.packages("vegan")} # useful ecology package
if(!require(ape)){install.packages("ape")} # useful ecology package
if(!require(lmerTest)){install.packages("lmerTest")} # useful package for statistics and longitudinal designs
if(!require(broom)){install.packages("broom")} #useful package for handling stats
if(!require(BiocManager)){install.packages("BiocManager")} # required for installing Bioconductor Packages
if(!require(devtools)){install.packages("devtools")} # required for installing GitHub Packages



# Packages from Bioconductor:
if(!require(dada2)){BiocManager::install("dada2")} # The original R package used by QIIME2
if(!require(ggtree)){BiocManager::install("ggtree")} # Plotting and manipulating trees
if(!require(philr)){BiocManager::install("philr")} # Testing of phylogenetic signals
if(!require(ShortRead)){BiocManager::install("ShortRead")}# load library(ShortRead)
if(!require(phyloseq)){BiocManager::install("phyloseq")}
if(!require(DESeq2)){BiocManager::install("DESeq2")}
if(!require(microbiome)){BiocManager::install("microbiome")}
if(!require(DECIPHER)){BiocManager::install("DECIPHER")}
if(!require(phangorn)){BiocManager::install("phangorn")}
if(!require(phangorn)){BiocManager::install("EnhancedVolcano")}
# Packages from GitHub
if(!require(qiime2R)){devtools::install_github("jbisanz/qiime2R")} # Package for reading artifacts
```




## Combine data into a phyloseq object
```{r 18,warning=F,message=F}
ps<-qza_to_phyloseq(
    features="table.qza",
    tree="rooted-tree.qza",
    "taxonomy.qza",
    metadata = "sample-metadata.txt"
    )
ps

```






### Filtering out non-Bacterial OTUs
```{r 18_1,warning=F,message=F}
ps1 <- ps %>%  subset_taxa(Kingdom == "Bacteria" & Family != "mitochondria" & Class != "Chloroplast")
ps1
```
```{r}
ps_melted <-  psmelt(ps1)
1 - sum(is.na(ps_melted$Genus))/nrow(ps_melted)
## [1] 0.726257
1 - sum(is.na(ps_melted$Species))/nrow(ps_melted)
## [1] 0.726257
```



### Plot OTUs at pylum level which has more than 1% aundance in each sample
```{r 18_1,warning=F,message=F}
ps1_phylum <- ps1 %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long forma
 filter(Abundance >= 0.01) %>%                         # Filter out low abundance taxa
  arrange(desc(Abundance)) 
# Set colors for plotting
phylum_colors <- c(
  "#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD",
   "#AD6F3B", "#673770","#D14285", "#652926", "#C84248", 
  "#8569D5", "#5E738F","#D1A33D", "#8A7C64", "#599861","#5F7FC7","#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD",
   "#AD6F3B", "#673770","#D14285","#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD", "#AD6F3B", "#673770","#CBD588"
)
# Plot 
Figure.phylum <- ggplot(ps1_phylum, aes(x = Subject, y = Abundance, 
                      fill = Phylum, 
                      label = sprintf("%1.0f%%", 100*round(Abundance, digits = 2)))) + 
  facet_wrap(~CariesRisk, scales="free_x", nrow=1) + 
  theme_bw()+
  geom_bar(stat = "identity") + 
  #geom_text(size = 1.7, position = position_stack(vjust = 0.5),aes(fontface=2),col = "white") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title = element_text(face = "bold", hjust = -0.2),
        axis.title.y = element_text(face = "bold",vjust = -4)) +
  scale_fill_manual(values = phylum_colors) + 
  scale_y_continuous(labels = function(x) paste0(x*100, "%")) +
  guides(fill = guide_legend(reverse = F, keywidth = 0.8, keyheight = 0.8)) +
  ylab("Relative Abundance (Phyla > 1%) \n") + xlab("Group by caries risk") +
  ggtitle("(a) At phylum level ") 

```


### Plot OTUs at Class level which has more than 1% aundance in each sample
```{r 18_2,warning=F,message=F}
ps1_class <- ps1 %>%
  tax_glom(taxrank = "Class") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  filter(Abundance >= 0.01) %>%                         # Filter out low abundance taxa
  arrange(desc(Abundance)) 
# Plot 
figure.class <- ggplot(ps1_class, aes(x = Subject, y = Abundance, 
                      fill = Class, 
                      label = sprintf("%1.0f%%", 100*round(Abundance, digits = 2)))) + 
  facet_wrap(~CariesRisk, scales="free_x", nrow=1) + 
  theme_bw()+
  geom_bar(stat = "identity") + 
  #geom_text(size = 1.7, position = position_stack(vjust = 0.5),aes(fontface=2),col = "white") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
         plot.title = element_text(face = "bold", hjust = -0.2),
        axis.title.y = element_text(face = "bold",vjust = -4)) +
  scale_fill_manual(values = phylum_colors) + 
  scale_y_continuous(labels = function(x) paste0(x*100, "%")) +
  guides(fill = guide_legend(reverse = F, keywidth = 0.8, keyheight = 0.8)) +
  ylab("Relative Abundance (Classes > 1%) \n") + xlab("Group by Caries Risk") +
  ggtitle("(b) At class level") 
```

### Plot OTUs at Family level which has more than 1% aundance in each sample
```{r 18_2,warning=F,message=F}
ps1_family <- ps1 %>%
  tax_glom(taxrank = "Family") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  filter(Abundance >= 0.01) %>%                         # Filter out low abundance taxa
  arrange(desc(Abundance)) 
# Plot 
figure.family <- ggplot(ps1_family, aes(x = Subject, y = Abundance, 
                      fill = Family, 
                      label = sprintf("%1.0f%%", 100*round(Abundance, digits = 2)))) + 
  facet_wrap(~CariesRisk, scales="free_x", nrow=1) + 
  theme_bw()+
  geom_bar(stat = "identity") + 
  #geom_text(size = 1.7, position = position_stack(vjust = 0.5),aes(fontface=2),col = "white") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title = element_text(face = "bold", hjust = -0.2),
        axis.title.y = element_text(face = "bold",vjust = -4)) +
  scale_fill_manual(values = phylum_colors) + 
  scale_y_continuous(labels = function(x) paste0(x*100, "%")) +
  guides(fill = guide_legend(reverse = F, keywidth = 0.8, keyheight = 0.8)) +
  ylab("Relative Abundance (Familiae > 1%) \n") + xlab("Group by Caries Risk") +
  ggtitle("(c) At family level") 
```







### Plot OTUs at Genus level which has more than 1% aundance in each sample
```{r 18_2,warning=F,message=F}
ps1_Genus <- ps1 %>%
  tax_glom(taxrank = "Genus") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  filter(Abundance >= 0.01) %>%                         # Filter out low abundance taxa
  arrange(desc(Genus)) 
# Plot 
figure.genus <- ggplot(ps1_Genus, aes(x = Subject, y = Abundance, 
                      fill = Genus, 
                      label = sprintf("%1.0f%%", 100*round(Abundance, digits = 2)))) + 
  facet_wrap(~CariesRisk, scales="free_x", nrow=1) + 
  theme_bw()+
  geom_bar(stat = "identity") + 
  #geom_text(size = 1.7, position = position_stack(vjust = 0.5),aes(fontface=2),col = "white") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title = element_text(face = "bold", hjust = -0.2),
        axis.title.y = element_text(face = "bold",vjust = -4)) +
  scale_fill_manual(values = phylum_colors) + 
  scale_y_continuous(labels = function(x) paste0(x*100, "%")) +
  guides(fill = guide_legend(reverse = F, keywidth = 0.8, keyheight = 0.8)) +
  ylab("Relative Abundance (Genera > 1%) \n") + xlab("Group by Caries Risk") +
  ggtitle("(d) At genus level")
```



##Generate taxa bar plot
```{r}
ggarrange(Figure.phylum,figure.class,figure.family,figure.genus, 
          ncol = 2, nrow = 2)

```



## alpha diversity

```{r 18,warning=F,message=F}
ps.rarefied = rarefy_even_depth(ps1, rngseed=1, sample.size=, replace=F)
plot_richness(ps.rarefied, x="CariesRisk", color  = "CariesRisk", measures = c("Chao1", "Shannon","Simpson")) +
  geom_boxplot() + ggtitle("Alpha Diversity Measure") + 
  theme_bw() + stat_compare_means(method = "anova", label = "p.signif",label.x = 2,size = 3.5,hide.ns = T) +
  stat_compare_means(method = "anova", size = 3, label.x  = 1) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        title = element_text(face = "bold"),
         text=element_text(size=15),
        legend.title = element_blank(),
        legend.position = "null") + 
  geom_point() + xlab("Caries Risk") + ylab("")
#stat_compare_means(comparisons = list(c("0ppm","10ppm"), c("10ppm","50ppm"), c("0ppm","50ppm"))) + stat_compare_means(label.y = 2.5) + xlab("")
```


## Ordination Plots
```{r  19, message=F, warning=F}
ps1.ord <- ordinate(ps1,"PCoA", "bray")
p2 = plot_ordination(ps1, ps1.ord,color = "CariesRisk") 
p2 + geom_point(size = 3) + ggtitle("(b) Bray Curtis") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0),
        title = element_text(face = "bold"),
        text=element_text(size=13),
        legend.title = element_blank())
```




```{r  19, message=F, warning=F}
ps1.ord <- ordinate(ps1,"PCoA", "unifrac")
p2 = plot_ordination(ps1, ps1.ord,color = "CariesRisk") 
p2 + geom_point(size = 3) + 
  ggtitle("(c) Unweighted UniFrac") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0),
        title = element_text(face = "bold"),
        text=element_text(size=13),
        legend.title = element_blank())
```



```{r  19, message=F, warning=F}
ps1.ord <- ordinate(ps1,"PCoA", "wunifrac")
p2 = plot_ordination(ps1, ps1.ord,color = "CariesRisk") 
p2 + geom_point(size = 3) + 
  ggtitle("(d) Weighted UniFrac") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0),
        title = element_text(face = "bold"),
        text=element_text(size=13),
        legend.title = element_blank())
```




########################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################
## Differential Abundace with DESeq2 


```{r  19, message=F, warning=F}

diagdds = phyloseq_to_deseq2(ps1, ~ CariesRisk)
gm_mean = function(x, na.rm=T){
  exp(sum(log(x[x > 0]), na.rm = na.rm) / length(x))
}
geoMeans = apply(counts(diagdds), 1, gm_mean)
diagdds = estimateSizeFactors(diagdds, geoMeans = geoMeans)
diagdds = DESeq(diagdds, fitType = "local")
res = results(diagdds)
res = res[order(res$padj, na.last=NA), ]
alpha = 0.05
sigtab = res[(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps1)[rownames(sigtab), ], "matrix"))

ggplot(sigtab, aes(x=Species, y=log2FoldChange, color=Genus)) + 
        geom_point(size=4) + coord_flip()
```







```{r  19, message=F, warning=F}
ps1.Genus = tax_glom(ps1, taxrank="Genus", NArm=FALSE)
diagdds = phyloseq_to_deseq2(ps1.Genus, ~ CariesRisk)
gm_mean = function(x, na.rm=T){
  exp(sum(log(x[x > 0]), na.rm = na.rm) / length(x))
}
geoMeans = apply(counts(diagdds), 1, gm_mean)
diagdds = estimateSizeFactors(diagdds, geoMeans = geoMeans)
diagdds = DESeq(diagdds, fitType = "local")
res = results(diagdds)
res = res[order(res$padj, na.last=NA), ]
alpha = 0.05
sigtab = res[(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps1.Genus)[rownames(sigtab), ], "matrix"))

ggplot(sigtab, aes(x=Genus, y=log2FoldChange, color=Family)) + 
        geom_point(size=4) + coord_flip()

```






```{r  19, message=F, warning=F}
ps1.Family = tax_glom(ps1, taxrank="Family", NArm=FALSE)
diagdds = phyloseq_to_deseq2(ps1.Family, ~ CariesRisk)
gm_mean = function(x, na.rm=T){
  exp(sum(log(x[x > 0]), na.rm = na.rm) / length(x))
}
geoMeans = apply(counts(diagdds), 1, gm_mean)
diagdds = estimateSizeFactors(diagdds, geoMeans = geoMeans)
diagdds = DESeq(diagdds, fitType = "local")
res = results(diagdds)
res = res[order(res$padj, na.last=NA), ]
alpha = 0.05
sigtab = res[(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps1.Family)[rownames(sigtab), ], "matrix"))
ggplot(sigtab, aes(x=Family, y=log2FoldChange, color=Order)) + 
        geom_point(size=4) + coord_flip()

```



```{r  19, message=F, warning=F}
ps1.Order = tax_glom(ps1, taxrank="Order", NArm=FALSE)
diagdds = phyloseq_to_deseq2(ps1.Order, ~ CariesRisk)
gm_mean = function(x, na.rm=T){
  exp(sum(log(x[x > 0]), na.rm = na.rm) / length(x))
}
geoMeans = apply(counts(diagdds), 1, gm_mean)
diagdds = estimateSizeFactors(diagdds, geoMeans = geoMeans)
diagdds = DESeq(diagdds, fitType = "local")
res = results(diagdds)
res = res[order(res$padj, na.last=NA), ]
alpha = 0.05
sigtab = res[(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps1.Order)[rownames(sigtab), ], "matrix"))
ggplot(sigtab, aes(x=Order, y=log2FoldChange, color=Class)) + 
        geom_point(size=4)+ coord_flip()

```



```{r  19, message=F, warning=F}
ps1.Class = tax_glom(ps1, taxrank="Class", NArm=FALSE)
diagdds = phyloseq_to_deseq2(ps1.Class, ~ CariesRisk)
gm_mean = function(x, na.rm=T){
  exp(sum(log(x[x > 0]), na.rm = na.rm) / length(x))
}
geoMeans = apply(counts(diagdds), 1, gm_mean)
diagdds = estimateSizeFactors(diagdds, geoMeans = geoMeans)
diagdds = DESeq(diagdds, fitType = "local")
res = results(diagdds)
res = res[order(res$padj, na.last=NA), ]
alpha = 0.05
sigtab = res[(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps1.Class)[rownames(sigtab), ], "matrix"))
ggplot(sigtab, aes(x=Class, y=log2FoldChange, color=Phylum)) + 
        geom_point(size=4) + coord_flip()

```



```{r  19, message=F, warning=F}
ps1.Phylum = tax_glom(ps1, taxrank="Phylum", NArm=FALSE)
diagdds = phyloseq_to_deseq2(ps1.Phylum, ~ CariesRisk)
gm_mean = function(x, na.rm=T){
  exp(sum(log(x[x > 0]), na.rm = na.rm) / length(x))
}
geoMeans = apply(counts(diagdds), 1, gm_mean)
diagdds = estimateSizeFactors(diagdds, geoMeans = geoMeans)
diagdds = DESeq(diagdds, fitType = "local")
res = results(diagdds)
res = res[order(res$padj, na.last=NA), ]
alpha = 0.05
sigtab = res[(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps1.Phylum)[rownames(sigtab), ], "matrix"))
ggplot(sigtab, aes(x=Phylum, y=log2FoldChange, color=Phylum)) + 
        geom_point(size=4) + coord_flip()

```








